Thank you, next: demographic consequences of partner diversity and turnover in a multi-species ant-plant mutualism

### 2024-09-23 Manuscript/ant_cactus_AmNat.tex
Updates
manuscript discussion

\begin{figure}
	\includegraphics[width=0.61\linewidth]{Figures/Lambdas_comp.png}
	\includegraphics[width=0.39\linewidth]{Figures/comp_num_partners.png}
	\caption{Panel a) shows the mean values of the estimated $\lambda_{NS}$ (filled in circles) and $\lambda_{S}$ (empty circles with an X) for each simulated combination of ant partners. The letters above the points correspond to what ant partners are present (V = Vacant, C = \textit{C. opuntiae}, L = \textit{L. apiculatum}, O = other). Panel b) shows the mean values of the estimated $\lambda_{NS}$ for }
	\label{fig:LambdaMeans}
\end{figure}

@@

\section*{Discussion}\ali{}{I am struggling reframing this part. I think I may need to start with a new outline, because a lot of what we had doesn't apply to our new conclusions that mutualism is good but we don't see a benefit of partner diversity?}
% mini abstract paragraph
Mutualisms commonly involve multiple partners but the ecological consequences of partner diversity remain poorly understood. 
Here we show that partner identity can play an important role in determining both vital rates and population fitness. 
The results of our heirarchical models revealed that different ant partners had different effects on vital rates, with \textit{C. opuniae} tended plants experiencing advantages in growth and survival when small, and \textit{L. apiculatum} tended plants experiencing floral viability advantages.
The results of our stochastic IPM revealed that all diversity scenarios which included partners resulted in the highest possible fitness for tree cholla, suggesting that partner diversity may not be beneficial or negative in this system. 
The results of our stochastic null IPM revealed that there is no evidence of portfolio effect in our system. 
These results highlight that partner presence can increase the overall benefits of a focal mutualism without synergistic benefits and the importance of a mechanistic understanding to explain the benefits of diversity across systems. \ali{}{This may not be great, but I am struggling a bit with this framing}

% Connect to broader patterns in the lit
Similar studies have reported complementarity \citep{Palmer2010,Rezende2007,Hooper2005,Cardinale2007,Afkhami2021} while we found that neither complementarity nor sampling effect explains the benefits of diversity in our system. 
In our first round of results, focused on individual ant effects on vital rates, we found that different ant partners affected different vital rates uniquely (Figures \ref{fig:Grow},\ref{fig:Surv},\ref{fig:Viab}).
This indicated to us that we may be dealing with complementarity, and that each ant partner offered a unique form of herbivore protection across ontogeny and or across different processes. 
However, our IPM results showed that we were actually observing a lack of beneficial diversity, leaving us with questions about how these results could fit together.

%This may have to do with the sheer number of \textit{L. apiculatum} in the system.
%None of these cited studies reported a single species with overwhelming frequency within their systems.
%In most cases of complementarity, there are relatively similar frequencies of each partner, which is not the case in our system. 
%75\% of observed ants are \textit{L. apiculatum} in our system, while only 17\% are \textit{C. opuntiae}.
%Our vital rate results suggested we may find evidence of complementarity, with \textit{C. opuntiae} appearing to boost growth and survial and \textit{L. apiculatum} appearing to boost floral viability.
%We believe there may simply not be enough \textit{C. opuntiae} in the system to have complementarity, leading instead to sampling effect driven by the overwhelming frequency of one ant.

% If there is one best partner then diversity should impose opportunity costs
%In systems where there is a single "best" partner and concurrent partner interactions, theory suggests that diversity should lead to more interactions with the best partner, thereby increasing the fitness of the focal mutualist \citep{Batstone2018}.
%In systems with mutually exclusive partner interactions and a "best`` partner,  theory suggests that diversity should impose an opportunity cost of the form of interacting with "inferior" partners, meaning scenarios where more than just the best partner is available should result in decreased fitness \citep{Noe1994}.
%While out system is aligned with this second expectation, we did not see any decreased fitness from the single "best" partner scenario to scenarios with the "best" partner and other ants included. 
%We do not see this reflected in our IPM results, potentially because of the high frequency of \textit{L. apiculatum}, the "best" partner, across the population.
%In the mutually exclusive interaction scenario, the fitness of the population should essentially be an average of the fitness conferred by each partner weighted by their frequency. 
%Due to the much higher frequency of \textit{L. apiculatum} ants, we believe the weighted average is almost equivalent to that of the single ant scenario with \textit{L. apiculatum} present.

%Turnover
Partner turnover is likely a significant driver behind the fitness we see within cholla populations, however the processes which drive the actual turnover frequency and directions remain a mystery in this system.
In the literature, it is clear that the frequency of partner turnover can have big effects on the fitness of the focal mutualist \cite{Fiala1994, Horvitz1986, Oliveira1999, Sachs2004}.
In some systems high freqency of turnover is necessary for resiliency and leads to higher fitness benefits \cite{Trojelsgaard2015}, while in other systems loyalty is the most beneficial \cite{Batstone2018}.
We found that some ants had higher frequency of turnover than others, specifically the other category appeared to have high frequency of turnover, rarely returning to the same plant, while \textit{C. opuntiae} and \textit{L. apiculatum} often returned.
The direction of partner turnover is also important when the identity of partners impacts the quality of benefits recieved \cite{Fonseca2003, Alonso1998, Dejean2008, Noe1994}.
In our study we found that there are distinct patterns to the direction of partner turnover. 
Specifically, most plants are tended by \textit{L. apiculatum} ants at some point, because vacant plants, other tended plants, and \textit{L. apiculatum} tended plants are all most likely to be tended by \textit{L. apiculatum} ants.
This indicates that while \textit{L. apiculatum} ants are loyal to their own plants and return multiple years in a row to the same ones, they are also strong colonizers.

What now remains is to determine the underlying processes that lead to the emergence of these characterized patterns. 
Throughout the literature, it is reported that non-symbiotic ants and non-mutualist ants are highly attracted to monosaccharide heavy EFN \citep{Heil2004,Kautz2009} and that plants often produce monosaccharide EFN to avoid exploitation \citep{Heil2010}.
It has been previously reported that when reproducing, the nectar of stem segments has a greater disaccharide vs monosaccharide ratio, while flowerbuds have a lower ratio \citep{Miller2008}. 
This could potentially have a significant impact on the partners who tend a plant in a given season, though determining the impacts of nectar on directional turnover requires further study. 
We conducted alternative simulations where the driving mechanism for ant occupation was not competitive exclusion, these results are included in Apendix XX.

@@

First, we were able to look at many years of data to reveal the complex dynamics of partner turnover and identity, revealing that the frequency of \textit{L. apiculatum} may be primarily responsible for the fitness of the cholla rather than any other ant partner observed.
Fewer years of data would not have been able to show the distinct patterns of \textit{L. apiculatum} taking over cacti that were previously tended by any other partners with as much clarity.
Second, our long-term data set allowed us to test the portfolio effect.
This data is required, because we must show that the partners react asynchronously to temporal environmental stochasticity for portfolio effect to even be possible.
This is a great example of how long-term data may offer more power to test underlying mechanisms in similar systems.
Finallly, previous studies by \cite{Miller2007} hypothesized that \textit{C. opuntiae} may be a parasite based on the effects on the cholla cacti.
Using more years of consecutive data in this system, we were able to show that \textit{C. opuntiae}, while the worst individual partner in terms of the overall cholla fitness (Figure \ref{fig:LambdaMeans}), increases the growth and survival rates of cacti prior to reproduction.
The length of our data set has allowed us to dive into individual vital rates as well as overall fitness with more clarity.

@@

\begin{figure}
\includegraphics[width=0.61\linewidth]{Figures/Lambdas_freq.png}
\includegraphics[width=0.39\linewidth]{Figures/freq_num_partners.png}
\caption{Panel a) shows the mean values of the estimated $\lambda_{S}$ (filled in circles) for each simulated combination of ant partners. The letters above the points correspond to what ant partners are present (V = Vacant, C = \textit{C. opuntiae}, L = \textit{L. apiculatum}, O = other). Panel b) shows the mean values of the estimated $\lambda_{S}$ for }
\label{fig:FreqLambdaMeans}
\end{figure}

\paragraph{Equal Likelihood Model}
The second alternative hypothesis we tested was what we called the equal likelihood model.
In this model we preserved the observed pattern of size-dependent vacancy/occupancy, but occupancy was manipulated to be equally likely for all partner identities. 
This was designed to remove the effect overwhelming numbers of \textit{L. apiculatum} ants may have. 
Despite very different proportions, we found very similar outcomes to the competitive exclusion model analyzed in the paper. 
All ants are beneficial, but having more than one is not necessarily any better than having an individual species as a partner (Figure \ref{fig:BetweenLambdaMeans}a).
Partner presence is beneficial, but neither identity nor number of partners appears to be important (Figure \ref{fig:BetweenLambdaMeans}b).

\begin{figure}
\includegraphics[width=0.61\linewidth]{Figures/Lambdas_equal.png}
\includegraphics[width=0.39\linewidth]{Figures/equal_num_partners.png}
\caption{Panel a) shows the mean values of the estimated $\lambda_{S}$ (filled in circles) for each simulated combination of ant partners. The letters above the points correspond to what ant partners are present (V = Vacant, C = \textit{C. opuntiae}, L = \textit{L. apiculatum}, O = other). Panel b) shows the mean values of the estimated $\lambda_{S}$ for }
\label{fig:BetweenLambdaMeans}
\end{figure}

>>>>

\begin{figure}
	\includegraphics[width=0.91\linewidth]{Figures/Lambdas_Comp_Full.png}
	\caption{Panel a) shows the mean values of the estimated $\lambda_{NS}$ (filled in circles) and $\lambda_{S}$ (empty circles with an X) for each simulated combination of ant partners. The letters above the points correspond to what ant partners are present (V = Vacant, C = \textit{C. opuntiae}, L = \textit{L. apiculatum}, O = other). Panel b) shows the mean values of the estimated $\lambda_{NS}$ for }
	\label{fig:LambdaMeans}
\end{figure}

@@

\section*{Discussion}
% mini abstract paragraph
Mutualisms commonly involve multiple partners but the ecological consequences of partner diversity remain poorly understood. 
Here we show that while partners can remain ecologically different, their identity can still be effectively equivalent.
The results of our heirarchical models revealed that different ant partners had different effects on vital rates, with \textit{C. opuniae} tended plants experiencing advantages in growth and survival when small, and \textit{L. apiculatum} tended plants experiencing floral viability advantages.
The results of our synchronous and non-synchronous stochastic IPM revealed that all diversity scenarios which included partners resulted in the highest possible fitness for tree cholla, suggesting that partner diversity is not beneficial in this system. 
These results highlight that partners can differ meaningfully while still resulting in the same fitness benefits for a focal mutualist.

% Connect to broader patterns in the lit
Similar studies have reported complementarity \citep{Palmer2010,Rezende2007,Hooper2005,Cardinale2007,Afkhami2021} while we found that there are no benefits of diversity in our system. 
In our first round of results, focused on individual ant effects on vital rates, we found that different ant partners affected different vital rates uniquely (Figures \ref{fig:Grow},\ref{fig:Surv},\ref{fig:Viab}).
This indicated to us that we may be dealing with complementarity, and that each ant partner offered a unique form of herbivore protection across ontogeny and or across different processes. 
However, our IPM results showed that we were actually observing a lack of beneficial diversity, leaving us with questions about how these results could fit together.

% why we don't see the vital rate differences in real life
Despite the apparent benefits for small \textit{C. opuntiae} tended cacti, these benefits are almost never experienced due to chronic vacancy of small cacti (Figure \ref{fig:Ant_Transition}).
We estimated that cacti between the approximate sizes of -$5 - 5 log(cm^3)$($0.01 - 148 cm^3$) would have important advantages in growth and survival if tended by \textit{C. opuntiae} ants.
At these sizes plants are very rarely tended ($< 10\%$ occupancy) due to lack of EFN production.
This means, while our vital rate models predict benefits for small ants, the assumptions we put in place for ant tending in our IPM means that these benefits are extremely rare and therefore have next to no impact on the fitness of the cacti.
This is not to say that growth and survival are not important vital rates, in fact they are likely to be the two most important in the model. 
In the literature, growth and survival for established plants are the vital rates that fitness estimates are most sensitive to \cite{Eacker2017}.
This indicates that the boosts recieved from ant tending of small plants would not be likely to make a large impact on fitness estimations even if this were allowed to occur in this model.

%Turnover
\ali{}{I could always reframe this to be We see little difference numerically based on ant partner identity, but ants have interesting turnover dynamics or something. Aka do I need to reshape this paragraph more?}
The processes which drive the actual turnover frequency and directions remain a mystery in this system.
In the literature, it is clear that the frequency of partner turnover can have big effects on the fitness of the focal mutualist \cite{Fiala1994, Horvitz1986, Oliveira1999, Sachs2004}.
In some systems high freqency of turnover is necessary for resiliency and leads to higher fitness benefits \cite{Trojelsgaard2015}, while in other systems loyalty is the most beneficial \cite{Batstone2018}.
We found that some ants had higher frequency of turnover than others, specifically the other category appeared to have high frequency of turnover, rarely returning to the same plant, while \textit{C. opuntiae} and \textit{L. apiculatum} often returned.
The direction of partner turnover is also important when the identity of partners impacts the quality of benefits recieved \cite{Fonseca2003, Alonso1998, Dejean2008, Noe1994}.
In our study we found that there are distinct patterns to the direction of partner turnover. 
Specifically, most plants are tended by \textit{L. apiculatum} ants at some point, because vacant plants, other tended plants, and \textit{L. apiculatum} tended plants are all most likely to be tended by \textit{L. apiculatum} ants.
This indicates that while \textit{L. apiculatum} ants are loyal to their own plants and return multiple years in a row to the same ones, they are also strong colonizers.

% rephrase this to be more relevant
What now remains is to determine the underlying processes that lead to the emergence of these characterized patterns. 
Throughout the literature, it is reported that non-symbiotic ants and non-mutualist ants are highly attracted to monosaccharide heavy EFN \citep{Heil2004,Kautz2009} and that plants often produce monosaccharide EFN to avoid exploitation \citep{Heil2010}.
It has been previously reported that when reproducing, the nectar of stem segments has a greater disaccharide vs monosaccharide ratio, while flowerbuds have a lower ratio \citep{Miller2008}. 
This could potentially have a significant impact on the partners who tend a plant in a given season, though determining the impacts of nectar on directional turnover requires further study. 
We conducted alternative simulations where the driving mechanism for ant occupation was not competitive exclusion, these results are included in Apendix C (Figures \ref{app:FreqLambdaMeans} - \ref{app:EqualLambdaMeans}).

@@

First, we were able to look at many years of data to reveal the complex dynamics of partner turnover and identity, revealing that despite the overwhelming  frequency of \textit{L. apiculatum} they do not impact the fitness of the cholla more than any other ant observed.
Fewer years of data would not have been able to show the distinct patterns of \textit{L. apiculatum} taking over cacti that were previously tended by any other partners with as much clarity.
Second, our long-term data set allowed us to test the portfolio effect.
This data is required, because we must show that the partners react asynchronously to temporal environmental stochasticity for portfolio effect to even be possible.
This is a great example of how long-term data may offer more power to test underlying mechanisms in similar systems.
Finallly, previous studies by \cite{Miller2007} hypothesized that \textit{C. opuntiae} may be a parasite based on the effects on the cholla cacti.
Using more years of consecutive data in this system, we were able to show that \textit{C. opuntiae}, while the worst individual partner in terms of the overall cholla fitness (Figure \ref{fig:LambdaMeans}), increases the growth and survival rates of cacti prior to reproduction.
The length of our data set has allowed us to dive into individual vital rates as well as overall fitness with more clarity.

@@

\begin{figure}
\includegraphics[width=0.61\linewidth]{Figures/Lambdas_freq.png}
\includegraphics[width=0.39\linewidth]{Figures/freq_num_partners.png}
\caption{Panel a) shows the mean values of the estimated $\lambda_{S}$ (filled in circles) for each simulated combination of ant partners. The letters above the points correspond to what ant partners are present (V = Vacant, C = \textit{C. opuntiae}, L = \textit{L. apiculatum}, O = other). Panel b) shows the mean values of the estimated $\lambda_{S}$ for }
\label{app:FreqLambdaMeans}
\end{figure}

\paragraph{Equal Likelihood Model}
The second alternative hypothesis we tested was what we called the equal likelihood model.
In this model we preserved the observed pattern of size-dependent vacancy/occupancy, but occupancy was manipulated to be equally likely for all partner identities. 
This was designed to remove the effect overwhelming numbers of \textit{L. apiculatum} ants may have. 
Despite very different proportions, we found very similar outcomes to the competitive exclusion model analyzed in the paper. 
All ants are beneficial, but having more than one is not necessarily any better than having an individual species as a partner (Figure \ref{fig:EqualLambdaMeans}a).
Partner presence is beneficial, but neither identity nor number of partners appears to be important (Figure \ref{fig:EqualLambdaMeans}b).

\begin{figure}
\includegraphics[width=0.61\linewidth]{Figures/Lambdas_equal.png}
\includegraphics[width=0.39\linewidth]{Figures/equal_num_partners.png}
\caption{Panel a) shows the mean values of the estimated $\lambda_{S}$ (filled in circles) for each simulated combination of ant partners. The letters above the points correspond to what ant partners are present (V = Vacant, C = \textit{C. opuntiae}, L = \textit{L. apiculatum}, O = other). Panel b) shows the mean values of the estimated $\lambda_{S}$ for }
\label{app:EqualLambdaMeans}
\end{figure}
